---
title: "When Funnel Plots Lie: Why Ecological Meta-Analyses Break Visual Diagnostics"
code-fold: true
reading-time: true
execute:
  warning: false
  message: false
filters: []
lang: en-GB
---

## Introduction 

Funnel plots are one of the most widely used visual tools in meta-analysis. They are typically presented as a simple diagnostic for small-study effects or publication bias: if the plot is symmetric, things are fine; if it is asymmetric, something is wrong. This interpretation is already fragile (see [this paper](https://doi.org/10.1016/j.jclinepi.2005.01.006)). In ecology, it is often **incorrect**.

Ecological meta-analyses rarely consist of independent, exchangeable study estimates. Instead, they are built from datasets with strong hierarchical structure: multiple effect sizes per study, shared controls, repeated measures, and outcomes that are intrinsically linked to study design and sample size.

When these data are plotted in a funnel plot, the resulting shape often reflects this structure rather than any underlying bias. In this post, I use simple simulations to show how **funnel plot asymmetry can arise without any publication bias at all**, and why visual interpretation is particularly misleading in ecological contexts.

## Setup 

```{r}
#| message: false
#| warning: false
#| 
library(tidyverse) 
library(metafor) 
library(patchwork) 
set.seed(123)

```

## A helper function for funnel plots

```{r}         

plot_funnel <- function(dat, title = "") {
  
  # meta-analytic model
  res <- rma(yi = yi, sei = sei, data = dat, method = "REML")
  mu <- as.numeric(res$b)
  
  sei_seq <- seq(min(dat$sei), max(dat$sei), length.out = 100)
  
  funnel_df <- tibble(
    sei = sei_seq,
    left = mu - 1.96 * sei_seq,
    right = mu + 1.96 * sei_seq
  )
  
  ggplot(dat, aes(x = yi, y = sei)) +
    geom_point(alpha = 0.6) +
    
    geom_line(data = funnel_df, aes(x = left, y = sei), linetype = "dashed") +
    geom_line(data = funnel_df, aes(x = right, y = sei), linetype = "dashed") +
    
    geom_vline(xintercept = mu) +
    
    scale_y_reverse() +
    labs(x = "Effect size", y = "Standard error", title = title) +
    theme_minimal()
}
```

## Scenario 1: The textbook case (independent studies)

We begin with the world that funnel plots assume: one independent effect size per study, with normally distributed sampling error.

```{r}         
n <- 40  
theta <- rnorm(n, 0, 0.5) 
sei <- runif(n, 0.1, 0.5) 
yi <- rnorm(n, theta, sei)  
dat1 <- tibble(yi, sei)  
p1 <- plot_funnel(dat1, "1. Independent studies (no bias)") 
p1
```

This is the idealised funnel: roughly symmetric, with increasing spread as precision decreases.

## Scenario 2: Classic publication bias

Now we introduce a simple selection mechanism: small, non-significant studies are less likely to be observed.

```{r}         
pvals <- 2 * pnorm(-abs(yi / sei))  
keep <- !(sei > 0.3 & pvals > 0.05)  
dat2 <- dat1[keep, ]  
p2 <- plot_funnel(dat2, "2. Publication bias (textbook case)") 
p2
```

This produces the familiar asymmetric funnel that is typically interpreted as evidence of bias.

## Scenario 3: Multiple effect sizes per study (no bias)

We now move into a more realistic ecological setting.

Each study contributes multiple effect sizes, sharing a common study-level effect.

```{r}         
n_study <- 40 
effects_per_study <- sample(3:8, n_study, replace = TRUE)  
dat3 <- map2_dfr(1:n_study, effects_per_study, function(j, k) {      u_j <- rnorm(1, 0, 0.5)  
sei_j <- runif(1, 0.1, 0.4)      
yi <- rnorm(k, u_j, sei_j)     
tibble(     yi = yi,     sei = rep(sei_j, k),     study = j   ) })  
p3 <- plot_funnel(dat3, "3. Hierarchical data (no bias)")
p3
```

Even without any publication bias, the funnel now shows clear structure: clustering, banding, and apparent asymmetry.

This arises purely from **non-independence**.

## Scenario 4: Shared controls and correlated outcomes

Now we simulate multiple effect sizes derived from the same underlying sample (e.g. shared controls or repeated measures).

```{r}         
dat4 <- map2_dfr(1:n_study, effects_per_study, function(j, k) {      u_j <- rnorm(1, 0, 0.5)   
base_se <- runif(1, 0.1, 0.3)      
yi <- rnorm(k, u_j, base_se)      
tibble(     yi = yi,     sei = rep(base_se, k),     study = j   ) })
p4 <- plot_funnel(dat4, "4. Shared controls / repeated measures") 
p4
```

The funnel now develops **bands and arcs**, which are often misinterpreted as evidence of bias but are in fact artefacts of shared variance structure.

## Scenario 5: Precision confounded with ecology (no bias)

In many ecological datasets, study size is not random. Smaller studies may focus on particular systems, species, or experimental designs.

Here, we simulate a case where effect size is associated with study size.

```{r}         
n <- 80  
sei <- runif(n, 0.1, 0.5)  
theta <- 0.8 * (sei > 0.3) + rnorm(n, 0, 0.2)  
yi <- rnorm(n, theta, sei)  
dat5 <- tibble(yi, sei)  
p5 <- plot_funnel(dat5, "5. Precision confounded with ecology") 
p5
```

This produces clear asymmetry — but there is no publication bias. The pattern arises because **small and large studies are studying different things**.

## Scenario 6: A realistic ecological meta-analysis

Finally, we combine these features:

-   multiple effect sizes per study

-   shared precision within studies

-   strong heterogeneity

-   confounding between study size and effect


```{r}         
dat6 <- map2_dfr(1:n_study, effects_per_study, function(j, k) {      sei_j <- runif(1, 0.1, 0.4)      
u_j <- rnorm(1, ifelse(sei_j > 0.25, 0.6, 0), 0.4)      
yi <- rnorm(k, u_j, sei_j)      
tibble(     yi = yi,     sei = rep(sei_j, k),     study = j   ) })  
p6 <- plot_funnel(dat6, "6. Realistic ecological structure") 
p6
```

This looks highly structured and asymmetric — yet no publication bias has been introduced.

## Putting it together

```{r}         
(p1 | p2) / (p3 | p4) / (p5 | p6)
```

## What these simulations show

Across these examples, the key point is consistent:

The funnel plot is not simply reflecting missing studies. It is reflecting the **structure of the data**.

In ecological meta-analyses, that structure often includes:

-    multiple non-independent effect sizes per study

-    shared controls or repeated measures

-    study-level clustering

-    strong heterogeneity

-    relationships between study design and precision
 

Under these conditions, the assumptions underlying funnel plot interpretation are violated.

## The problem with visual interpretation

The usual advice is to “inspect the funnel plot for asymmetry.” However, this assumes that asymmetry primarily arises from small-study effects.

These simulations show that:

-   asymmetry can arise without any bias

-   structured patterns (bands, arcs, clusters) are common

-   visually striking funnels can be entirely artefactual


In this context, visual interpretation is not just subjective, it is often **misleading**.

## What should we do instead?

Rather than relying on visual inspection of funnel plots, analyses should:

-    explicitly account for hierarchical structure

-    consider study-level aggregation as a sensitivity check

-    use multilevel meta-analytic models

-    explore moderators that may be confounded with precision

-    interpret any small-study effect diagnostics with extreme caution


## Take-home message

Funnel plots were designed for a world of independent study estimates.

Ecological meta-analyses do not live in that world.

Before interpreting a funnel plot, we should first ask:


> What structure do these points represent?

In many cases, the answer is not “a set of independent studies,” but a complex, hierarchical evidence base and the funnel plot is simply revealing that structure.


## A better way to explore small-study effects in ecology

If ecologists want to explore whether effect sizes are associated with study precision, a more defensible approach is to use a **multilevel meta-regression** rather than relying on visual inspection of a raw funnel plot. This follows the general approach advocated by [Nakagawa et al.](https://doi.org/10.1111/2041-210X.13724), who recommend modelling heterogeneity and non-independence directly, and then using **residual funnel plots** as a supplement rather than a primary diagnostic. 

The idea is simple: if smaller or less precise studies tend to report systematically different effects, then effect size may be associated with sampling variance or standard error. In ecological meta-analysis, however, this association should be tested in a multilevel framework that reflects the dependency structure of the data.

```{r}
#| label: "Simulate data and assess small-study effects"
#| message: false
#| warning: false

library(dplyr)
library(purrr)
library(tibble)
library(metafor)
library(ggplot2)

set.seed(123)

# Assume your data look something like this:
# dat$yi        = effect size
# dat$vi        = sampling variance
# dat$study_id  = study identifier
# dat$obs_id    = effect-size identifier within study
# dat$year      = publication year (optional, for time-lag bias)

simulate_meta_ecology <- function(
  n_study = 40,
  min_es = 3,
  max_es = 8,
  tau = 0.5,              # between-study heterogeneity
  sigma_within = 0.2,     # within-study variation
  bias = FALSE,           # simulate publication bias
  confounding = FALSE,    # precision-effect confounding
  time_lag = FALSE        # time trend
) {
  
  # number of effect sizes per study
  n_es <- sample(min_es:max_es, n_study, replace = TRUE)
  
  dat <- map2_dfr(seq_len(n_study), n_es, function(j, k) {
    
    # study-level precision (shared within study)
    sei_j <- runif(1, 0.1, 0.5)
    vi_j  <- sei_j^2
    
    # study-level true effect
    if (confounding) {
      # less precise studies tend to have larger true effects
      mu_j <- rnorm(1, mean = ifelse(sei_j > 0.3, 0.6, 0), sd = tau)
    } else {
      mu_j <- rnorm(1, mean = 0, sd = tau)
    }
    
    # optional time trend
    year_j <- sample(1990:2020, 1)
    if (time_lag) {
      mu_j <- mu_j + 0.02 * (year_j - 2000)
    }
    
    # generate observed effect sizes
    yi <- rnorm(k, mean = mu_j, sd = sqrt(vi_j + sigma_within^2))
    
    tibble(
      yi = yi,
      vi = rep(vi_j, k),
      study_id = factor(j),
      obs_id = factor(paste0(j, "_", seq_len(k))),
      year = rep(year_j, k)
    )
  })
  
  # optional publication bias / small-study selection
  if (bias) {
    dat <- dat %>%
      mutate(
        sei = sqrt(vi),
        z = yi / sei,
        p = 2 * pnorm(-abs(z))
      ) %>%
      filter(!(sei > 0.3 & p > 0.05)) %>%
      select(-z, -p)
  }
  
  dat %>%
    mutate(sei = sqrt(vi))
}

# Simulate a dataset:
# - confounding = TRUE gives a precision-effect relationship without publication bias
# - time_lag = TRUE adds a publication year trend
dat <- simulate_meta_ecology(
  n_study = 40,
  min_es = 3,
  max_es = 8,
  tau = 0.5,
  sigma_within = 0.2,
  bias = FALSE,
  confounding = TRUE,
  time_lag = TRUE
)

# ---------------------------------------------------------
# Multilevel meta-analysis (baseline model)
# ---------------------------------------------------------

m0 <- rma.mv(
  yi = yi,
  V  = vi,
  random = ~ 1 | study_id/obs_id,
  data = dat,
  method = "REML"
)

summary(m0)

# ---------------------------------------------------------
# Multilevel small-study effect model
# (Egger-type meta-regression, but multilevel)
# ---------------------------------------------------------

m_pub <- rma.mv(
  yi = yi,
  V  = vi,
  mods = ~ sei,
  random = ~ 1 | study_id/obs_id,
  data = dat,
  method = "REML"
)

summary(m_pub)

# The coefficient for 'sei' is the key term:
# if it differs clearly from zero, effect size changes with precision,
# which is consistent with a small-study effect
# (but not necessarily publication bias)

# ---------------------------------------------------------
# Time-lag bias model
# ---------------------------------------------------------

m_time <- rma.mv(
  yi = yi,
  V  = vi,
  mods = ~ scale(year),
  random = ~ 1 | study_id/obs_id,
  data = dat,
  method = "REML"
)

summary(m_time)

# ---------------------------------------------------------
# Residual funnel plot
# ---------------------------------------------------------

dat$resid_pub <- as.numeric(residuals(m_pub))

mu_resid <- 0
sei_seq <- seq(min(dat$sei), max(dat$sei), length.out = 200)

funnel_df <- tibble(
  sei = sei_seq,
  left = mu_resid - 1.96 * sei_seq,
  right = mu_resid + 1.96 * sei_seq
)

ggplot(dat, aes(x = resid_pub, y = sei)) +
  geom_point(alpha = 0.6) +
  geom_line(data = funnel_df, aes(x = left, y = sei), linetype = "dashed") +
  geom_line(data = funnel_df, aes(x = right, y = sei), linetype = "dashed") +
  geom_vline(xintercept = 0) +
  scale_y_reverse() +
  labs(
    x = "Residual effect size",
    y = "Standard error",
    title = "Residual funnel plot after multilevel meta-regression"
  ) +
  theme_minimal()
```


This is not a magic solution. No publication-bias method is ideal in the kinds of hierarchical, heterogeneous datasets common in ecology, and [Nakagawa et al.](https://doi.org/10.1111/2041-210X.13724) are very clear on that point. A multilevel meta-regression plus a residual funnel plot is much closer to the structure of ecological evidence than a visually interpreted raw funnel plot.
